Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-03-01 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Feedback on introduction section

\Abstract{
This is a tutorial paper on imputing incomplete multilevel data with
\pkg{mice}. Footnotes in the current version show work in progress/under
construction. The last section is not part of the manuscript, but purely
for reminders.
}

@@

In many contemporary data analysis efforts, some form of hierarchical or
clustered structure is recorded. For example, students may be clustered
in classes in psychometrics research, or patients may be clustered in
studies in individual patient data meta-analyses. Ignoring the clustered
structure of such multilevel data can be harmful to the statistical
inferences and bias results. Imagine a case where cross-level
interactions between unit-level variables and cluster-level variables
are present. The cluster to which a unit belongs may then influence the
unit-level observations--and vice versa for each of the units that make
up the cluster \citep{hox17}. These relations can and should be taken
into account when developing analysis models for multilevel data for the
simple reason that groups of observations share some common variance.
The variability due to clustering is often measured by means of the
intraclass coefficient \citep[ICC,][]{fish25}. The ICC can be seen as
the percentage of variance that can be attributed to the cluster-level,
where a high ICC would indicate that a lot of variability is due to the
cluster structure. Multilevel models typically accommodate for this
variability by including a separate group mean for each cluster. In
addition to random intercepts, multilevel models can also include random
effects and heterogeneous residual error variances across clusters
\citep[see e.g.][\citet{hox17} and \citet{jong21}]{gelm06}. There are
many names for models that take clustering into account. Some popular
examples are `multilevel models', `hierarchical models', `mixed effect
models' and `random effect models'.

@@

{\centering \includegraphics{Manuscript_files/figure-latex/patterns-1}

@@

The process of analyzing multilevel data is further complicated when not
all data entries are observed. Just as with single level data,
missingness may occur at the unit level. But with multiple levels of
data comes the potential for clustered missingness. Incomplete
multilevel data can be categorized into two general patterns: systematic
missingness and sporadic missingness \citep{resc13}. Systematic
missingness implies that one or more variables are never observed in a
certain cluster. With sporadic missingness there may be observed data
for some but not all units in a cluster \citep{buur18, jola18}. We have
visualized this difference in Figure 1, which shows an \(n \times p\)
set \(\mathbf{X} = X_1, \dots, X_p\), with \(n\) units distributed over
\(N\) clusters, and \(p=3\) columns. Column \(X1\) is completely
observed, column \(X2\) is systematically missing, and column \(X3\) is
sporadically missing. To analyze these incomplete data, we have to take
the nature of the missingness and the cluster structure into account.
For example, the sporadic missingness in \(X3\) could be easily amended
if this would be a cluster-level variable (and thus constant within
clusters). We could then just extrapolate the true (but missing) value
of \(X3\) for unit 1 from unit 2, and the value for unit 4 from unit 3.
If \(X3\) would instead be a unit-level variable (which may vary within
clusters), we could not just recover the unobserved `truth', but need
some kind of missing data method or discard the incomplete units
altogether (i.e., list-wise deletion/complete case analysis). And, with
the systematic missingness in \(X2\), it is impossible to fit a
multilevel model since we cannot estimate the intercept of cluster 2. We
would have to exclude this cluster from our analyses entirely to obtain
any results. Obviously, excluding observations is not a desirable
workflow.

Ignoring the missingness in analyses can be extremely harmful to
inferences. Complete case analysis can introduce bias in statistical
inferences and lowers statistical power. Instead, the missingness should
be accommodated \emph{before} or \emph{within} the analysis of
scientific interest. Especially the former is very generic and popular.
Imputing (i.e., filling in) the missing values separates the missing
data problem from the scientific problem: missing data are replaced by
plausible values whereafter the completed data is analyzed as if it were
completely observed. The \proglang{R} package \pkg{mice} has become the
de-facto standard for imputation by chained equations, which iteratively
solves the missingness on a variable-by-variable basis. \pkg{mice} is
known to yield valid inferences under many different missing data
circumstances \citep{buur18}. In this paper, we will discuss how to use
\pkg{mice} in the context of multilevel data.

@@

\begin{itemize}
\tightlist
\item
  \texttt{popmis} from the \pkg{mice} package (simulated data on
  perceived popularity, \(n = 2,000\) pupils across \(N = 100\)
  schools);
\item
  \texttt{hiv} from the \pkg{GJRM} package (simulated data on HIV
  diagnoses, \(n = 6,416\) patients across \(N = 9\) regions);
\item
  \texttt{impact} from the \pkg{metamisc} package (empirical data on
  traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
  studies).
\end{itemize}

For each of these datasets, we will discuss the nature of the
missingness, choose one or more imputation models and evaluate the
imputed data, but we will also highlight one specific aspect of the
imputation workflow. With the \texttt{popmis} data, we show how (not) to
develop an imputation model. With the \texttt{hiv} data we focus on
extending the imputation model to include Heckman-type
selection-inclusion methods. And with the \texttt{impact} data we
provide an example of multivariate missingness in real-world data.
Together, this should give enough scaffolding for applied researchers
who are faced with incomplete multilevel data.

@@

@@

{\centering \includegraphics{Manuscript_files/figure-latex/pop_pat-1}

@@

{\centering \includegraphics{Manuscript_files/figure-latex/pop_dist-1}

@@

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_ignored_eval-1} \end{center}

@@

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_predictor_eval-1} \end{center}

@@

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_norm_eval-1} \end{center}

@@

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_pan_eval-1} \end{center}

@@

\begin{center}\includegraphics{Manuscript_files/figure-latex/hiv-1} \end{center}

@@

\begin{itemize}
\tightlist
\item
  \texttt{name} Name of the study,
\item
  \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
  observational cohort),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{motor\_score} Glasgow Coma Scale motor score,
\item
  \texttt{pupil} Pupillary reactivity,
\item
  \texttt{ct} Marshall Computerized Tomography classification,
\item
  \texttt{hypox} Hypoxia (0=no, 1=yes),
\item
  \texttt{hypots} Hypotension (0=no, 1=yes),
\item
  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
\item
  \texttt{edh} Epidural hematoma (0=no, 1=yes),
\item
  \texttt{mort} 6-month mortality (0=alive, 1=dead).
\end{itemize}

The analysis model for this dataset is a prediction model with
\texttt{mort} as the outcome.\footnote{Look at analysis model, maybe
  copy from GREAT data example e.g., adjusted prognostic effect of ct on
  unfortunate outcomes. Add something about systematically missing data
  here.} The data is already imputed (Steyerberg et al, 2008), so we've
induced missingness again based on the missingness in the original data.

@@

\begin{center}\includegraphics{Manuscript_files/figure-latex/impact-1} \end{center}

@@

\begin{center}\includegraphics{Manuscript_files/figure-latex/unnamed-chunk-3-1} \end{center}

\begin{center}\includegraphics{Manuscript_files/figure-latex/unnamed-chunk-3-2} \end{center}

@@

>>>>

\Abstract{
This is a tutorial paper on imputing incomplete multilevel data with
\pkg{mice}. Footnotes in the current version show work in progress/under
construction. The last section is not part of the manuscript, but purely
for reminders. We aim to submit at JSS, so there is no word count limit
(``There is no page limit, nor a limit on the number of figures or
tables'').
}

@@

\sout{In many contemporary data analysis efforts, some form of
hierarchical or clustered structure is recorded.} Contemporary decision
making is driven increasingly often and sometimes even entirely dictated
by data. Although data relevant for decision makers were initially
collected using small and well-designed studies, there has been a
growing need to address more complex questions at a wider scale. This
not only requires to collect larger amounts of data, but also from more
diverse settings and populations. For example, {[}TD: introduce IPDMA
here; eg. HIV studies Johanna is referring to{]}. Another example is the
use of large databases with information from thousands or even millions
of individuals from multiple cities, regions or even countries. Finally,
a third situation arises when data are collected from multicenter
studies {[}TD: maybe refer to the schools/classes example here{]}. For
example, students may be clustered in classes in psychometrics research,
or patients may be clustered in studies in individual patient data
meta-analyses in medical research. A common characteristic in
aforementioned examples is the presence of clustering. {[}TD: briefly
explain what is clustering?{]}

Ignoring the clustered structure of such multilevel data can be harmful
to the statistical inferences and introduce bias in estimators
\citep{hox17}. \sout{Imagine a case where cross-level interactions
between unit-level variables and cluster-level variables are present.
The cluster to which a unit belongs may then influence the unit-level
observations--and vice versa for each of the units that make up the
cluster.} These relations can and should be taken into account when
developing analysis models for multilevel data for the simple reason
that groups of observations share some common variance. \sout{The
variability due to clustering is often measured by means of the
intraclass coefficient (ICC). The ICC can be seen as the percentage of
variance that can be attributed to the cluster-level, where a high ICC
would indicate that a lot of variability is due to the cluster
structure.} Multilevel models typically accommodate for this variability
by including a separate group mean for each cluster. In addition to
random intercepts, multilevel models can also include random effects and
heterogeneous residual error variances across clusters \citep[see
e.g.][\citet{hox17} and \citet{jong21}]{gelm06}. There are many names
for models that take clustering into account. Some popular examples are
`multilevel models', `hierarchical models', `mixed effect models' and
`random effect models'. {[}TD: emphasize when/why we need to account for
clustering in the analysis of clustered data. Why is the presence of
clustering relevant when considering multiple imputation of missing
data? e.g.~distinction between systematically and sporadically missing
data. But also: mechanisms of missing data (e.g.~MCAR, MAR, MNAR) may
differ between clusters. But also: relation between observed data may
differ between clusters? When/ why should we avoid using traditional
imputation methods? e.g.~congeniality issues.{]}

@@

{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/patterns-1}

@@

The process of analyzing multilevel data is further complicated when not
all data entries are observed. Just as with single level data,
missingness may occur at the unit level. But with multiple levels of
data comes the potential for clustered missingness. Therefore,
incomplete multilevel data can be categorized into two general patterns:
systematic missingness and sporadic missingness \citep{resc13}.
Systematic missingness implies that one or more variables are never
observed in a certain cluster. With sporadic missingness there may be
observed data for some but not all units in a cluster
\citep{buur18, jola18}. We have visualized this difference in Figure 1,
which shows an \(n \times p\) set \(\mathbf{X} = X_1, \dots, X_p\), with
\(n\) units distributed over \(N\) clusters, and \(p=3\) columns. Column
\(X1\) is completely observed, column \(X2\) is systematically missing
in cluster 2, and column \(X3\) is sporadically missing. To analyze
these incomplete data, we have to take the nature of the missingness and
the cluster structure into account. For example, the sporadic
missingness in \(X3\) could be easily amended if this would be a
cluster-level variable (and thus constant within clusters). We could
then just extrapolate the true (but missing) value of \(X3\) for unit 1
from unit 2, and the value for unit 4 from unit 3. If \(X3\) would
instead be a unit-level variable (which may vary within clusters), we
could not just recover the unobserved `truth', but would need to use
some kind of missing data method or discard the incomplete units
altogether (i.e., list-wise deletion/complete case analysis). Further,
with the systematic missingness in \(X2\), it is impossible to fit a
multilevel model since we cannot estimate the intercept of cluster 2. We
would have to exclude this cluster from our analyses entirely to obtain
any results. Obviously, excluding observations is not a desirable
workflow.

Ignoring the missingness in analyses can be extremely harmful to
inferences. Complete case analysis can introduce bias in statistical
inferences and lowers statistical power. Instead, the missingness should
be accommodated \emph{before} or \emph{within} the analysis of
scientific interest. Especially the former is very generic and popular.
{[}VJ: add why multiple values are imputed.{]} Imputing (i.e., filling
in) the missing values separates the missing data problem from the
scientific problem: missing data are replaced by plausible values
whereafter the completed data is analyzed as if it were completely
observed. The \proglang{R} package \pkg{mice} has become the de-facto
standard for imputation by chained equations, which iteratively solves
the missingness on a variable-by-variable basis. \pkg{mice} is known to
yield valid inferences under many different missing data circumstances
\citep{buur18}. In this paper, we will discuss how to use \pkg{mice} in
the context of multilevel data.

@@

\begin{itemize}
\tightlist
\item
  \texttt{popmis} from the \pkg{mice} package \citep[simulated data on
  perceived popularity, \(n = 2,000\) pupils across \(N = 100\)
  schools,][]{mice};
\item
  \texttt{hiv} from the \pkg{GJRM} package \citep[simulated data on HIV
  diagnoses, \(n = 6,416\) patients across \(N = 9\) regions,][]{GJRM};
\item
  \texttt{impact} from the \pkg{metamisc} package \citep[empirical data
  on traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
  studies,][]{metamisc}.
\end{itemize}

For each of these datasets, we will discuss the nature of the
missingness, choose one or more imputation models and evaluate the
imputed data, but we will also highlight one specific aspect of the
imputation workflow. With the \texttt{popmis} data, we show how (and how
not) to develop an imputation model. With the \texttt{hiv} data we focus
on extending the imputation model to include Heckman-type
selection-inclusion methods. With the \texttt{impact} data we provide an
example of multivariate missingness in real-world data. Together, this
should give enough scaffolding for applied researchers who are faced
with incomplete multilevel data.

@@

TODO: make a `translation table' between model equation and lme4. Use
betas instead of gamma's and mu's.

@@

{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_pat-1}

@@

{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_dist-1}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_ignored_eval-1} \end{center}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_predictor_eval-1} \end{center}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_norm_eval-1} \end{center}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_pan_eval-1} \end{center}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/hiv-1} \end{center}

@@

\begin{itemize}
\tightlist
\item
  \texttt{name} Name of the study,
\item
  \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
  observational cohort),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{motor\_score} Glasgow Coma Scale motor score,
\item
  \texttt{pupil} Pupillary reactivity,
\item
  \texttt{ct} Marshall Computerized Tomography classification,
\item
  \texttt{mort} 6-month mortality (0=alive, 1=dead).
\end{itemize}

The analysis model for this dataset is a prediction model with
\texttt{mort} as the outcome.\footnote{Look at analysis model, maybe
  copy from GREAT data example e.g., adjusted prognostic effect of
  \texttt{ct} on unfortunate outcomes, we just want to know the adjusted
  odds ratio for \texttt{ct}. Add something about systematically missing
  data here.} The data is already imputed (Steyerberg et al, 2008), so
we've induced missingness again based on the missingness in the original
data.

\[
\text{mort}_{ij} =
\gamma_{00} + 
\gamma_{01} \text{ type}_{j} + 
\gamma_{10} \text{ age}_{ij} + 
\gamma_{10} \text{ moter\_score}_{ij} + 
\gamma_{10} \text{ pupil}_{ij} + 
\gamma_{10} \text{ ct}_{ij} + \\
u_{0j} + 
u_{1j} \text{ age}_{ij} + 
u_{1j} \text{ moter\_score}_{ij} +  
u_{1j} \text{ pupil}_{ij} +  
u_{1j} \text{ ct}_{ij} +  
e_{ij} \\
\]

\texttt{glmer(mort\ \textasciitilde{}\ 1\ +\ type\ +\ age\ +\ motor\_score\ +\ pupil\ +\ ct\ +\ hypox\ +\ hypots\ +\ tsah\ +\ edh\ +\ (1\ \textbar{}\ name))}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-1} \end{center}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact_md-1} \end{center}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact_md-2} \end{center}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> mod_cca <- glmer(mort ~ type + age + motor_score + pupil + ct + (1 | name), family = "binomial", data = impact_NA) 
\end{CodeInput}
\begin{CodeOutput}
Warning in checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv, : Model is nearly unidentifiable: very large eigenvalue
 - Rescale variables?
\end{CodeOutput}
\begin{CodeInput}
R> mod_cca %>% broom.mixed::tidy()
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 9 x 7
  effect   group term            estimate std.error statistic   p.value
  <chr>    <chr> <chr>              <dbl>     <dbl>     <dbl>     <dbl>
1 fixed    <NA>  (Intercept)      -2.66     0.194      -13.7   1.11e-42
2 fixed    <NA>  typeRCT          -0.412    0.172       -2.40  1.64e- 2
3 fixed    <NA>  age               0.0298   0.00190     15.7   2.56e-55
4 fixed    <NA>  motor_score3     -0.631    0.0823      -7.66  1.82e-14
5 fixed    <NA>  motor_score4     -0.994    0.0845     -11.8   5.61e-32
6 fixed    <NA>  motor_score5/6   -1.54     0.0868     -17.7   2.88e-70
7 fixed    <NA>  pupil             0.438    0.0397      11.0   3.01e-28
8 fixed    <NA>  ct                0.430    0.0343      12.5   5.34e-36
9 ran_pars name  sd__(Intercept)   0.257   NA           NA    NA       
\end{CodeOutput}
\begin{CodeInput}
R> # use glm with name as dummy
R> # only random effect on ct, not the other variables
R> # make forest plot of ct variable
R> # is there heterogen in association of ct with mort?
R> # is there any random effect on that variable at all?
R> # 
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- quickpred(impact_NA)
R> pred[pred == 1] <- 2
R> pred[, "name"] <- -2
R> pred["mort", ] <- 2
R> pred[, "mort"] <- 2
R> diag(pred) <- 0
R> pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
R> pred
\end{CodeInput}
\begin{CodeOutput}
            name type age motor_score pupil ct hypox hypots tsah edh mort
name           0    0   0           0     0  0     0      0    0   0    0
type           0    0   0           0     0  0     0      0    0   0    0
age            0    0   0           0     0  0     0      0    0   0    0
motor_score    0    0   0           0     0  0     0      0    0   0    0
pupil         -2    0   0           2     0  2     0      2    0   0    2
ct            -2    2   2           2     2  0     0      0    2   2    2
hypox         -2    2   0           2     0  0     0      2    0   0    2
hypots        -2    0   0           2     2  0     2      0    0   0    2
tsah          -2    0   2           0     0  2     0      0    0   0    2
edh           -2    0   0           0     0  2     0      0    0   0    2
mort           0    0   0           0     0  0     0      0    0   0    0
\end{CodeOutput}
\begin{CodeInput}
R> meth <- make.method(impact_NA)
R> meth[meth != ""] <- "2l.pmm"
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> # imp_impact <- impact_NA %>%
R> #   mutate(name = as.integer(name), motor_score = as.numeric(motor_score)) %>%
R> #   mice::mice(., m = 2, maxit = 1, method = meth, predictorMatrix = pred)
R> # look at jomo for categorical variables?
R> # semi-cont with jomo is not ideal (schafer, '97) because you need 2-step approach
R> # pmm is better (more efficient) because it will still look for donors (maybe outside of cluster) based on predictive distance, even for very small clusters
R> # make assumptions of these methods explicit!
\end{CodeInput}
\end{CodeChunk}
